---
title: "Scientific graphics with `ggplot2` in R"
author: "Jan K Boyer"
date: "10/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

##Why Code your plots?
- Integrate analysis and figure creation
- *Reproducible*
- *Free and open source*
- *Excellent documentation* The official ggplot2 documentation is excellent, and numerous online help forums and blog posts provide even more resources for troubleshooting and learning.
- *Protects us from ourselves* `ggplot2` makes it difficult or impossible to add problematic features (3d, dual axes), forcing us to make clean, scientific graphs.


# R Graphics Packages

There are several ways to make graphs in R. R comes with graphics capability (referred to as base graphics), and packages such as `lattice` and `ggplot2` can also be used to make graphics. 

**figure out how to do columns**
## base

```{r base_example_plot}
plot(Sepal.Length~Petal.Width, data = iris)
```

## `lattice`

```{r lattice_example_plot}
require(lattice)
xyplot(Sepal.Length~Petal.Width|Species, data = iris)
```

## `ggplot2`
```{r echo=FALSE}
theme_set(theme_grey(base_size = 20))
```

```{r ggplot_example_plot}
require(ggplot2)
ggplot(iris, aes(x = Sepal.Length, y = Petal.Width)) +
  geom_point()
```

This course focuses on `ggplot2`, which is the package I use most often. However, you can also make excellent publication quality graphics with base graphics or `lattice`.

# Formatting data

Before making a `ggplot2', make sure your data is in the right format:

1. All data in one dataframe
2. Data in long format  
    - Each observation is a row
    - Each variable is a column
3. Column classes correct
4. Factors ordered correctly

This section provides code examples for the data formatting edits that are most commonly needed to correctly format data for plotting. All code blocks show both base R and tidyverse code, so you can use whichever you are most comfortable with. 

## 1. All data in one dataframe

All of your data needs to be in one dataframe before making a ggplot. If you have different dataframes for each site, species, or replicate, or have site information in one dataframe and species data in another, you need to combine your dataframes into one before plotting. 

### Add observations (rows)
If you have multiple dataframes with identical column names, you can simple paste bind the rows together:


```{r eval=FALSE}
#base
#columns must match
df.all <- rbind(df1, df2)

#dplyr (tidyverse)
#can handle missing columns
require(dplyr)
df.all <- bind_rows(df1, df2)
```


### Add variables (columns)


####add exampled df

```{r eval=FALSE}
#base
df.new <- merge(df1, df2, by = "siteID") #inner
df.new <- merge(df1, df2, by = "siteID", all.x = TRUE) #left outer
df.new <- merge(df1, df2, by = "siteID", all.y = TRUE) #right outer
df.new <- merge(df1, df2, by = "siteID", all = TRUE) #outer

#dplyr (tidyverse)
#will automatically join on common column names
#use by = "siteID" to specify join variable
df.new <- df1 %>% inner_join(df2) #inner
df.new <- df1 %>% left_join(df2) #left outer
df.new <- df1 %>% right_join(df2) #right outer
df.new <- df1 %>% full_join(df2) #outer
```

### Convert wide data to long data

We often record data in wide format

.pull-left[
# Wide
```{r echo=FALSE}
d.wide <- tibble(
  date = rep(as.Date('2019-06-23') + c(0,0,0,2,2,2)),
  site = rep(c("a", "b", "c"), 2),
  FMS = rpois(6, 23),
  BHS = rpois(6, 9), 
  HBC = rpois(6, 2) 
)

d.long <- gather(d.wide, #data
                 key = species,  #new key column (i.e., what are groups)
                 value = catch,  #new value column (i.e., what did you measure)
                 -date, -site) #columns that should not be gathered


knitr::kable(d.wide, format = 'html')
```
]

.pull-right[
#Long
```{r echo=FALSE}
knitr::kable(d.long, format = 'html')
```
]


##Convert wide to long

```{r wide_long_conversion, eval=FALSE}
#tidyverse
require(tidyr)

gather(data, key = "key", value = "value")
```
  


*Other options:  
`datatable::melt()`   
`pivot_longer()` in development version of `tidyverse`*



d.wide <- tibble(
  date = rep(as.Date('2019-06-23') + c(0,0,0,1,1,1,2,2,2)),
  site = rep(c("a", "b", "c"), 3),
  flannelmouth = rpois(9, 23),
  bluehead = rpois(9, 9), 
  carp = rpois(9, 2) 
)

d.long <- gather(d.wide, #data
                 key = species,  #new key column (i.e., what are groups)
                 value = n_caught,  #new value column (i.e., what did you measure)
                 -date, -site) #columns that should not be gathered



### Check column classes
```{r eval=FALSE}
str(iris) #base r
glimpse(iris) #tidyverse
```

### Define column classes
```{r eval=FALSE}
df$weight <- as.numeric(df$weight) #base

df <- df %>% mutate(weight = as.numeric(weight)) #dplyr (tidyverse)
  
as.numeric()
as.integer()
as.character()
as.factor()
as.Date() #format should be 2019-04-30
as.POSIXct() #format: 2019-04-30 14:22:00
```


# Order Factors

In order to arrange character or factor data in a certain order, you must make it a factor and define levels. Otherwise, `ggplot2` will simply arrange it alphabetically.

```{r }
#base
iris$Species <- factor(iris$Species,
                       levels = c("virginica", "versicolor", "setosa")) #<<

#dplyr (tidyverse)
iris <- iris %>% 
  mutate(Species = factor(Species, 
                          levels = c("virginica", "versicolor", "setosa"))) #<<

levels(iris$Species) #check they are correct now
```


# Making ggplots

ggplots are made out of the following elements:

### `data`
### `aes`
### `geom`
### `stat`
### `coord`
### `scale`
### `facet` 
### `theme` 

Some of the above elements are optional, but there are three elements that you *always* need to define for a ggplot:

### `data`
### `aes`thetic mappings map variables to x, y, color, shape, etc.
### `geom`etric objects display data on plot (point, line, boxplot, etc.)

This animation shows how we add these elements to our code to make a simple ggplot:
ADD GIF

### A shortcut to prettier plots:

The default theme for `ggplot2`, `theme_gray()` isn't great. 

```{r echo = FALSE, fig.width=8, fig.height=6}
mtcars$gear <- factor(mtcars$gear)
ggplot(mtcars, aes(x = wt, y = mpg, colour = gear)) +
  geom_point()
```

ggplot2 comes with several built in themes that are much cleaner than `theme_gray()`

```{r echo = FALSE, out.width = "100%", fig.asp = 0.5}
require(cowplot)
mtcars$gear <- factor(mtcars$gear)
p <- ggplot(mtcars, aes(x = wt, y = mpg, colour = gear)) +
  geom_point()

classic <-  p +  theme_classic(base_size = 10) +
  theme(plot.title = element_text(family = "mono")) + 
  ggtitle("theme_classic()")
minimal <- p +  theme_minimal(base_size = 10)+
  theme(plot.title = element_text(family = "mono")) +
  ggtitle("theme_minimal()")
          
bw <- p +  theme_bw(base_size = 10)+
  theme(plot.title = element_text(family = "mono")) + 
  ggtitle("theme_bw()")
light <- p + theme_light(base_size = 10) +
  theme(plot.title = element_text(family = "mono")) +
  ggtitle("theme_light()")

plot_grid(classic,  minimal, bw, light, nrow = 2)

```
[See all built in themes](https://ggplot2.tidyverse.org/reference/index.html#section-themes)

The `theme_set()` function sets a default theme. Add `theme_set()' with your preferred theme and font size to the top of your R script, and then all of the graphs you create in that script will use that theme.

```{r out.width = "70%", fig.height=3}
theme_set(theme_minimal(base_size = 20)) #<<

ggplot(mtcars, aes(x = wt, y = mpg, color = gear)) +
  geom_point()
```
